---
title: "Intro to R: The Basics"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
  pdf_document:
    toc: TRUE
    toc_float: TRUE
  html_notebook:
    toc: TRUE
    code_folding: "none"
editor_options: 
  markdown: 
    wrap: sentence
---

In an R Markdown document, you can type and **format** text, include images, as 
well as execute code in the console or in code chunks, as the one below.

To run a line of code using Windows (or Linux), put the cursor on the same line 
then press **ctrl + enter**. The same shortcut works to run a highlighted part 
of code. To run the entire code chunk press **ctrl + shift + enter**.
On macOS, replace **cmd** for **ctrl** to run a line of code, a highlighted part 
of code or to run the entire code chunk. For the latter, you can also use the 
functionally equivalent shortcut **cmd + option + c**.

### Our first chunk

```{r first chunk}
1 + 1 

#This is a comment

2 + 2 #this is also a comment

#you can comment or un-comment a line by highlighting it 
#then pressing ctrl + shift + c

"this is not a comment"


```

**Tip: Add comments to your code, describing what you are doing and why you are doing it.**
Comments are an easy way to document what you are doing to other people (friends, coworkers, supervisors, ...) and future you (to remember what you were doing and why you were doing it)

For more details on how to format and use R Markdown, either consult the [official website](https://rmarkdown.rstudio.com/lesson-1.html) or the [cheatsheet](https://rmarkdown.rstudio.com/lesson-15.html). No need to save the 
link: In RStudio, you can always access it under 
**Help \> Cheat Sheets \> R Markdown Cheat Sheet**.

An overview of all shortcuts (for your current operating system) is available
under **Tools > Keyboard Shortcuts Help**.


### R as calculator

```{r calculator}
5 + 2 #Addition 
5 - 2 #Subtraction
5 * 2 #Multiplication
5 / 2 #Division
5^2   #Exponent
1+3*5 #mixed
(1+2)+2^2*3 #can you guess the answer?

```

But R is more than just a calculator.
You can carry out more complex operations using *functions*.

### Functions

```{r functions}
#get the square root of 4
4^(0.5)

#or using the function: sqrt()
sqrt(4)

#get the sum of 2 + 3
2 + 3

#or using the function: sum()
sum(2,3)

#find max value in a series of numbers 
max(1,2,3,4)

#or minimum
min(1,2,3,4)

#round a number
round(3.142)

#get the absolute value of a number: |x|
abs(-3)
abs(3)

#you can print a text
print("Functions make life easier") 

```

Feel free to try some operations and use some functions yourself. 
You can **add a new code chunk** or run code in the console below. 
- To clear the console, click at a new line on the console then press ctrl + l.
- To add a new code chunk, click the *Insert Chunk* button on the toolbar above
or by pressing *Ctrl+Alt+I* on Windows or *Cmd+Option+I* on macOS.

**ADD NEW CODE CHUNK HERE**

You can also add a new code chunk yourself by clicking the *Insert Chunk* button on the toolbar or by pressing **Ctrl+Alt+I** on Windows or **Cmd+Option+I** on macOS.

### Functions with arguments

So the form of a function is the name of the function followed by ( ).
What goes inside the parentheses is called the *arguments* of the function.
So far, we only provided the function with an input argument, but further arguments can be provided as extra options for how the function would work.

```{r arguments}
#remember the rounding task?
round(3.142)

#it is equivalent to
round(3.142, digits = 0)


#what if we want rounding to the 1 or 2 decimal place?
round(3.142, digits = 1)
round(3.142, digits = 2)


#if you are more familiar with the arguments of the function
#you can skip providing the name of the argument
round(3.142, 1)
round(3.142, digits = 1)
round(x = 3.142, digits = 1)


#to get more information about a function and its arguments
#type ? then the name of the function

?round

#functions can take other functions as input
sqrt(2)           #answer is approx. 1.414214
round(sqrt(2), 1) #answer is 1.4

#can make it even more complicated, r works from inside out
round(sqrt(sum(1,1)), 1)

#notice where every function starts and ends
#by clicking after a bracket to see the bracket closing it 
round(   
  sqrt(
    sum(1,1)
             )
              , 1)

#did you notice that r does not care about spaces and new lines?


#lets look at the paste() function
paste("Welcome", "to", "Sharjah")
paste("Welcome", "to", "Sharjah", sep = " ")

#what is the time?
paste("11", "45", sep = ":")

#what happens if you remove the "sep ="?
paste("11", "45", ":")


#let's add seconds
paste("11", "45", "12", sep = ":")


#be creative
paste("Peter Parker", "MJ", sep = " *hearts* ")

```

Let us take some time to play with functions.

### Some exercises (part 1)

**Problem 1**

You want to start an online platform to fight misinformation.
You expect the project to start financing itself after **10** years, so you get a loan from the bank for **1.9** million euros with an interest rate of **2.5%** for the entire period.

How much do you expect to have paid the bank by the end of the period?

```{r}
#you can use this following formula:
#total amount = the principal * (1 + interest rate * years)




#round the amount to the first and second decimal place



#hint: first answer is 2.38 millions, second is 2.4 millions
```

**Problem 2**

A voter has to choose between two candidates.
She has a position of **5** on the cultural left-right scale and **2** on the economic left-right scale.
She puts a weight of **0.7** on cultural issues and **0.3** on economic issues.
Candidate A has position of **6** on both cultural and economic issues, while candidate B has position of **3** on both cultural and economic issues.
Which candidate is she more likely to vote for?

```{r}
#hint: she is likely to vote for the candidate with the smaller distance
#to her issue positions, use these formulas:
# D_A = W_E * |P_EV - P_EA| + W_C *|P_CV - P_CA|
# D_B = W_E * |P_EV - P_EB| + W_C *|P_CV - P_CB|

#where:
#D_i is distance of voter to Candidate i
#W_E is the weight the voter puts on economic issues
#W_C is the weight the voter puts on cultural issues
#P_Ei is the position of person i on the economic scale
#P_Ci is the position of person i on the cultural scale










###hint: D_A = 1.9, D_B = 1.7

```

As you must have noticed in the last example.
The formula for choosing the smallest distance could get quite complicated.

```{r}
min(0.3 * abs(2-6) + 0.7 *abs(5-6), 0.3 * abs(2-3) + 0.7 *abs(5-3))

#this would have been easier

D_A <- 0.3 * abs(2 - 6) + 0.7 *abs(5 - 6)
D_B <- 0.3 * abs(2 - 3) + 0.7 *abs(5 - 3)

min_D <- min(D_A, D_B)


```

What did we do here?
We stored the distances in an **object**.
This is a key advantage of using R (and other programming languages), the ability to store the results of your calculations for further manipulations.
You can think of objects as storage containers for your calculations.

### Objects

```{r objects}
#you can assign an object using the <- operator

age <- 23
income <- 1400
field <- "political science"
university <- "University of Sharjah"


#you can also use the = operator but it is not recommended
age = 23

#stored objects appear in the environment
#to free up memory, you can remove objects using rm() command
rm(age)

#you can remove all objects using the command
rm(list = ls())


```

There are rules for naming objects.

-   Objects can contain numbers but must start with a character: **2014GDP** is invalid while **GDP2014** is valid.

-   Object names should not contain a blank space, use an underscore "**\_**" instead.
    For example: **vote_share** instead of **vote share**.

-   Avoid using other special characters such as **? ! \*** and so on.

-   Object names in R are case sensitive, so object *Age* is different from *age*.

-   Avoid using names of functions such as sum, mean, and print...etc for your objects.

    A good coding practice is not to **hard code** your values.
    lets look at an example in the following chunk:

```{r}
#if you are calculating the area of a trianlge with base length 2 cm and height 3 cm, you can go about it like this:

tri_area <- 0.5 * 2 * 3

#but a "cleaner" way of doing it would be:
base <- 2
height <- 3
tri_area <- 0.5 * base * height

#this way you don't have to manually change the "2" and "3" everywhere if you need to use these dimensions again later in the code



```

Objects can store different *data types*:

1.  Numeric, basically numbers such as 0, 1, 1.2, -32.1, and -3.21e1
2.  Character, text such as "vote", "SPD", "1". Character input should always be provided between single ' ' or double " " quotation marks, otherwise R will think your input is referring to the name of an object that does not exist.
3.  Logical, which is true or false statement. In the syntax of R: **TRUE** or **T** and **FALSE** or **F**.
4.  Factors, which is similar to character but is used to indicate a category, such as "Republican", "Democrat", "Independent".
5.  There are other data types that are less frequently used, such as a Date format, but we will focus on the main ones here.

You can check the type of an object using the `class()` command, you can also convert the type of an object (where possible) with commands such as `as.numeric` .
Let's have a look below:

```{r data types}

area <- 15.1

gdp_growth  <- -1.2

country <- "UAE"

grade <- "1.3"

Sharjah_is_fun <- TRUE #notice the color of TRUE

climate_doing_well <- F  #F or T as shorthand for FALSE or TRUE


#Let's check the classes 

class(gdp_growth)
class(country)
class(climate_doing_well)


#let's convert the class of grade into numeric
class(grade)

grade <- as.numeric(grade)

class(grade)

#you can convert grade into a character with as.character()
grade <- as.character(grade)

#expect an error message if you try to convert country into numeric


#TRUE and FALSE are treated like 1 and 0
number <- 1
number <- number + TRUE
number <- number * FALSE

```

If some data is not acting as you are expecting, there is a chance the type is not what you want it to be, e.g., you are processing the data as numeric but what you have is character or factor.
Just keep an eye on the class of your data.

Very well, now take few minutes to create your own objects and assign values to them.
Check their classes and try to convert between data types.
See what works and what does not.

```{r}









```

### Vectors

Almost always we don't work with one single data point, but our dataset has many observations.
For example, the age of one person is not enough to establish the relationship between one's age and vote preferences.
We need a series of age observations.
A series of observation in R is called a **vector**.
You can use the command `c()` to create a vector from your input.

```{r vectors}

philosophers <- c("Hegel", "Nietzsche", "Leibniz", "Kant")

print(philosophers)

age <- c(33, 40, 32, 34, 31, 26, 29)

print(age)

revolution_success <- c(TRUE, FALSE, FALSE, TRUE, TRUE)

print(revolution_success)

#you can the number of elements in a vector using length()
n_age <- length(age)

#we can use our functions on vectors
mean_age <- sum(age)/length(age)

#or easier
mean_age <- mean(age)


#same as before with min and max
min_age <- min(age)
max_age <- max(age)


#you can perform mathematical operations on vectors 
revenue <- c(100, 80, 120, 150)
cost <- c(30, 10, 20, 40)
profit <- revenue - cost

print(profit)


prices <- c(80, 100, 40, 160)
prices_inflation <- prices * 1.25
print(prices)
print(prices_inflation)


#what happens when you mix numeric and character values in a vector?
confused_vector <- c(5, 2, "summer", "winter")
print(confused_vector)

#the numeric values turn into character


```

### Some more exercises (part 2)

**Task 1**

```{r Task 1}
#think about 5 cities you visited or want to visit
#write their names in a vector


#try to guess the temperature now in each city
#write these temperatures in a vector


#multiply these temperatures by 2 then save it in a new vector


#create a logical vector that takes the value TRUE if humans can survive this new temperature and FALSE if they can't

```

**Task 2**

```{r Task 2}
#In some political surveys, the left-right political ideology scale is from 0 to 10, 0 being extreme left and 10 being extreme right
#guess the ideological placement of the following voters and create a vector of the values:
#Christian Democrat, the Left party, Liberal party, Green Party, AFD, and Social Democrat



#modify your vector so the ideological placement would be scaled from 0 to 1

```

**Task 3**

```{r Task 3}
#In some political surveys, political interest is scaled from 1-4, where 1 means very interested in politics, and 4 means very uninterested in politics. This is counter intuitive. create a vector for the political interest values of the following survey participants **then** reverse the scale so 4 would mean very interested in politics

#participant 1: very interested
#participant 2: somewhat uninterested
#participant 3: very uninterested
#participant 4: somewhat interested
#participant 5: somewhat uninterested
#participant 6: very interested

pi_old <- c(1,3,4,2,3,1)



#hint: use the formula: new_value = |old_value -(max(old_value) + min (old_value))|

```


Before we move on from vectors, let us look at creating a vector from a sequence 
of numbers and selecting specific elements from vectors.

```{r}

#fast ways to create a sequence of numbers 
first_ten_numbers <- c(1:10)
print(first_ten_numbers)

first_ten_numbers <- seq(from = 1, to = 10, by = 1)
print(first_ten_numbers)

first_ten_numbers <- seq(1, 10, 1) #because we know the order of the arguments
print(first_ten_numbers)

first_odd_numbers <- seq(1, 10, 2) 
print(first_odd_numbers)

first_even_numbers <- seq(2, 10, 2)
print(first_even_numbers)


#selecting elements from a vector
#let's create a vector with the English alphabet

alphabet <- c("A", "B", "C", "D", "E")

#you can select an element from a vector by providing
#the index (location) of the element in the vector between []
first_letter <- alphabet[1]
print(first_letter)

second_letter <- alphabet[2]
print(second_letter)

#same for the rest of the elements

#remember the length() command, it lets find the last element
length(alphabet) #gives number of elements in the vector
last_letter <- alphabet[length(alphabet)]

#selecting multiple elements
first_three_letters <- alphabet[1:3] #1:3 is similar to c(1,2,3)

print(first_three_letters) #we created a subset of our vector

first_three_letters <- alphabet[c(1,2,3)]

print(first_three_letters)

#how about selecting everything except the first element
all_but_first <- alphabet[2:length(alphabet)]
print(all_but_first)

#even better
all_but_first <- alphabet[-1]
print(all_but_first)

#negative indices means drop that element


#logical values can be used to subset vectors
first_three_letters <- alphabet[c(T,T,T,F,F)]

#TRUE means to include this element, FALSE means to remove it
#we will explore the true potential of this after some exercises


```



We mentioned that you can select from/subset vectors using logical vectors.
But we do not need to manually specify which elements to include or exclude, 
we can define certain *conditions* for including elements. For example, if you 
have a vector of student grades, you can choose to include only students above or
below a certain grade. To define these conditions, we first need to get to know 
some logical operators.

### Logical Operators

You can use R also for TRUE/FALSE statements for logical statements via the comparison operators

-   `>` greater than
-   `<` smaller than
-   `==` equal
-   `!=` not equal
-   `>=` greater than or equal to
-   `<=` less than or equal to
-   `%in%` equal to at least one element in a group

```{r}


1 < 3 

5 >= 8 

11 != 10  

22 == 22  

7 < 3  

5 <= 2+3 

5 %in% c(2,4,5,7,8)

```

You can also use logical operators:

-   `&` element-wise AND operator. It returns TRUE if both elements are true
-   `|` element-wise OR operator. It returns TRUE if one of the statements is TRUE
-   `!` Logical NOT operator. It returns FALSE if statement is TRUE

```{r}

5 & 4 < 8 

#equivalent to
5 < 8 & 4 < 8


5 | 4 < 8 

#equivalent to
5 > 8 | 4 < 8 


! 5 > 2 

#try some of these conditions in the console


```

Now let us apply conditions to subset vectors.

```{r}

#we have a vector for temperatures
temp <- c(20, 41, 10, -2 , 35, -3, 8, 18)

temp == 20 # find room temperature
room_temp <- temp[temp == 20]
print(room_temp)


temp > 30 #find very hot temperatures
hot_temp <- temp[temp > 30]
print(hot_temp)


temp < 0 #find freezing temperatures
freezing_temp <- temp[temp < 0]
print(freezing_temp)


temp > 0 & temp < 15 #find cold temperatures
cold_temp <- temp[temp > 0 & temp < 15 ]
print(cold_temp)


temp > 15 & temp < 30 #find nice temperatures
nice_temp <- temp[temp > 15 & temp < 30]
print(nice_temp)



```

### Some more exercises (part 3)

**Rising populist threat in Democracies:**

In a potential not so far future, left and right wing populists in Democracies are gaining support across electoral districts.
Assume a country that has about 300 electoral districts, we generate fictional data for these 300 electoral districts.

The goal of the task is to select and count the following districts:

Safe democratic: vote populist \< 25%

significant populist minority: 25% \<= vote populist \< 33%

blocking populist minority: 33% \<= vote populist \< 50% ( can block decisions requiring two-thirds majority)

absolute populist majority: 50% \<= vote populist \< 63%

super populist majority: vote populist \>= 63%

**Hint**: you can count the number of elements in a vector with the `length()` function

```{r}

set.seed(2024) 
vote_populist <- rnorm(300, 40, 12) 
#we generate 300 normally distributed random values using rnorm()
#set.seed() makes our random draws reproducible




```



