```{r setup, message=FALSE, warning=FALSE, results='hide'}

p_needed <-
  c("tidyverse", "haven", "MASS", "sjPlot")

packages <- rownames(installed.packages())

p_to_install <- p_needed[!(p_needed %in% packages)]

if (length(p_to_install) > 0) {
  install.packages(p_to_install)
}

sapply(p_needed, require, character.only = TRUE)


```


```{r}

d1 <- read.csv("issp_2016_jp.csv")

m1 <- lm(inc_jp ~ edu, data = d1)
m2 <- lm(inc_jp ~ edu + female, data = d1)

```


```{r}


# plot(jitter(d1$edu, factor = 1), jitter(d1$inc_jp, factor = 1),
#      col = ifelse(d1$female ==1, alpha("blue",0.4),
#                   alpha("red",0.4)), pch = 19 ,
#      ylab = "Income", xlab = "Years of Education")

plot(jitter(d1$edu, factor = 1), jitter(d1$inc_jp, factor = 1),
     col = alpha("black",0.4), pch = 19 ,
     ylab = "Income", xlab = "Years of Education")

abline(a = coef(m1)[1], b = coef(m1)[2], lwd = 3)

abline(a = coef(m2)[1], b = coef(m2)[2], lwd = 3, col = "red")
abline(a = coef(m2)[1] + coef(m2)[3], b = coef(m2)[2], lwd = 3, col = "blue")


legend("topleft", c("Male", "Female"),
        pch = 19, col = c("red", "blue"), cex = 0.95)



```




## Step 1 - Get the regression coefficients.

```{r step-1-get-the-regression-coefficients}
beta_hat <- coef(m2)
```

## Step 2 - Generate sampling distribution. 

### Step 2.1. Get the variance-covariance matrix.  

```{r step-2-1-get-the-variance-covariance-matrix}
V_hat <- vcov(m2) 

```

### Step 2.2. Draw from the multivariate normal distribution. 

```{r step-2-2-draw-from-the-multivariate-normal-distribution}

# Set the number of draws/simulations.

nsim <- 1000 

# Draw from the multivariate normal distribution to get S.

S <- mvrnorm(nsim, beta_hat, V_hat)

dim(S) # Check dimensions

```

## Step 3 - Set a scenario. 

Tip: double-check the ordering of coefficients first, to put the values in the correct order.

```{r step-3-set-a-scenario}
names(beta_hat)
edu_seq <- c(8:20)  
scenario_female <- cbind(1, edu_seq, 1) 
scenario_male <- cbind(1, edu_seq, 0)

```
 
## Step 4 - Calculate Expected Values 

```{r expected-values}

#EV_female <- S %*% as.matrix(X_female)  

EV_female <- S %*% t(scenario_female)  

# %*% is the operator for matrix multiplication

EV_male <- S %*% t(scenario_male)

quants_mean_fun <-  function(x) {
  c(quants = quantile(x, probs = c(0.025, 0.975)),
    mean = mean(x))
}

# Quantiles, we again use apply and our quants.mean.fun
quants_female <- apply(EV_female, 2, quants_mean_fun)
quants_male <- apply(EV_male, 2, quants_mean_fun)


```



## Step 5 - Summarize Results

```{r summarize-result-base-r-2}
plot(
  d1$edu,
  d1$inc_jp,
  pch = 19,
  cex.axis = 0.8,
  cex.lab = 0.8,
  bty = "n",
  las = 1,
 # ylim = c(0, 35),
  ylab = "Income",
  xlab = "Years of Education",
  type = "n"
)

points(jitter(d1$edu, factor = 1), jitter(d1$inc_jp, factor = 1),
      col = ifelse(d1$female ==1, alpha("blue",0.4),
                   alpha("red",0.4)), pch = 19 ,
      ylab = "Income", xlab = "Years of Education")

polygon(
  x = c(edu_seq, rev(edu_seq)),
  y = c(quants_female[1 ,], rev(quants_female[2 ,])),
  col =  alpha("blue",0.2),
  border = NA
)
polygon(
  x = c(edu_seq, rev(edu_seq)),
  y = c(quants_male[1 ,], rev(quants_male[2 ,])),
  col = alpha("red",0.2),
  border = NA
)


lines(edu_seq, quants_female[3,], col = "blue", lwd = 2)

lines(edu_seq, quants_male[3,],
     col = "red", lwd = 2)

# Add a legend

legend("topleft", c("Male", "Female"),
        pch = 19, col = c("red", "blue"), cex = 0.95)


```




###using fake data for a more equal Japan


```{r}


#create more equal income

d1$inc_jp_equal <- ifelse(d1$female == 1, d1$inc_jp + 1900, d1$inc_jp)


m3 <- lm(inc_jp_equal ~ edu + female, data = d1)

```



```{r simulations equal}
beta_hat <- coef(m3)

V_hat <- vcov(m3) 

nsim <- 1000 

S <- mvrnorm(nsim, beta_hat, V_hat)

dim(S) # Check dimensions

edu_seq <- c(8:20)  
scenario_female <- cbind(1, edu_seq, 1) 
scenario_male <- cbind(1, edu_seq, 0)


EV_female <- S %*% t(scenario_female)  
EV_male <- S %*% t(scenario_male)

quants_female <- apply(EV_female, 2, quants_mean_fun)
quants_male <- apply(EV_male, 2, quants_mean_fun)


```



```{r results equal}
plot(
  d1$edu,
  d1$inc_jp_equal,
  pch = 19,
  cex.axis = 0.8,
  cex.lab = 0.8,
  bty = "n",
  las = 1,
 # ylim = c(0, 7000),
  ylab = "Income",
  xlab = "Years of Education",
  type = "n"
)

 points(jitter(d1$edu, factor = 1), jitter(d1$inc_jp_equal, factor = 1),
       col = ifelse(d1$female ==1, alpha("blue",0.4),
                    alpha("red",0.4)), pch = 19 ,
       ylab = "Income", xlab = "Years of Education")

polygon(
  x = c(edu_seq, rev(edu_seq)),
  y = c(quants_female[1 ,], rev(quants_female[2 ,])),
  col =  alpha("blue",0.2),
  border = NA
)
polygon(
  x = c(edu_seq, rev(edu_seq)),
  y = c(quants_male[1 ,], rev(quants_male[2 ,])),
  col = alpha("red",0.2),
  border = NA
)


lines(edu_seq, quants_female[3,], col = "blue", lwd = 2)

lines(edu_seq, quants_male[3,],
     col = "red", lwd = 2)

# Add a legend

legend("topleft", c("Male", "Female"),
        pch = 19, col = c("red", "blue"), cex = 0.95)


```

